Effectful Semantics for Substructural SSA

### 2023-11-07 pldi24/paper.tex
Work on trace monad section, citations

% Syntax
\newcommand{\letexpr}[3]{\ensuremath{\ms{let}\;#1 = #2\;\ms{in}\;#3}}
\newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
\newcommand{\brb}[2]{\ms{br}\;#1\;#2}
\newcommand{\lbrb}[2]{\brb{\lbl{#1}}{#2}}
\newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
\newcommand{\ewhere}[2]{\ms{then}\;#1\;\ms{where}\;#2}
\newcommand{\where}[2]{#1\;\ms{where}\;#2}
\newcommand{\wbranch}[3]{#1(#2) \lto #3}
\newcommand{\lwbranch}[3]{\wbranch{\lbl{#1}}{#2}{#3}}
\newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
%\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
\newcommand{\csplits}[3]{#1 \mapsto #2;#3}
\newcommand{\cwk}[2]{#1 \mapsto #2}
\newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
\newcommand{\tlin}[2]{#2 \subseteq \ms{lin}(#1)}
\newcommand{\ltlin}[3]{#3 \subseteq \ms{lin}(#1) \cap #2}
\newcommand{\thyp}[3]{#1: {#2}^{#3}}
\newcommand{\lhyp}[3]{#1[#2](#3)}
\newcommand{\llhyp}[3]{\lhyp{\lbl{#1}}{#2}{#3}}
\newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
\newcommand{\taff}{{\{\ms{a}\}}}
\newcommand{\trel}{{\{\ms{r}\}}}
\newcommand{\tint}{{\{\ms{a}, \ms{r}\}}}
\newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
\newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\issubst}[3]{#1: #2 \mapsto #3}
\newcommand{\lbsubst}[3]{#1: #2 \rightsquigarrow #3}
\newcommand{\exprletsubst}[2]{{#1};{#2}}
\newcommand{\stmtletsubst}[2]{{#1};{#2}}
\newcommand{\mhole}[1]{{#1}^?}
\newcommand{\lhole}[1]{?#1}
\newcommand{\mhasty}[6]{#1;#2 \vdash_{#3} #4: {#5}^{#6}}
\newcommand{\mhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\mlhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\tyhole}[5]{#1: #2 \mapsto_{#3} {#4}^{#5}}
\newcommand{\blkhole}[3]{#1: #2 \mapsto #3}
\newcommand{\cfghole}[3]{#1: #2 \mapsto #3}
\newcommand{\substctx}[2]{{#1}^{#2}}
\newcommand{\substlbs}[2]{{#1}^{#2}}
\newcommand{\restrictsubst}[2]{{#1}_{#2}}
\newcommand{\subsubst}[2]{#1 \subseteq #2}
\newcommand{\isrw}[3]{#1: #2 \mapsto #3}
% \newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}

@@

@@

@@

\TODO{Ye Olde Trace Monad}

\TODO{Is Elgot}

\TODO{Ye Olde Nondeterministic Trace Monad}

\TODO{Is Elgot}

\TODO{Ye Olde Reader/Writer/State Transformer}

\TODO{Is Elgot}

\TODO{already done above?}

@@

\TODO{\cite{sparky}}

\TODO{Pomsets...}

@@

We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{String}^* + \ms{String}^\omega)\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.

@@

@@

>>>>

% Syntax
\newcommand{\letexpr}[3]{\ensuremath{\ms{let}\;#1 = #2\;\ms{in}\;#3}}
\newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
\newcommand{\brb}[2]{\ms{br}\;#1\;#2}
\newcommand{\lbrb}[2]{\brb{\lbl{#1}}{#2}}
\newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
\newcommand{\ewhere}[2]{\ms{then}\;#1\;\ms{where}\;#2}
\newcommand{\where}[2]{#1\;\ms{where}\;#2}
\newcommand{\wbranch}[3]{#1(#2) \lto #3}
\newcommand{\lwbranch}[3]{\wbranch{\lbl{#1}}{#2}{#3}}
\newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
%\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
\newcommand{\csplits}[3]{#1 \mapsto #2;#3}
\newcommand{\cwk}[2]{#1 \mapsto #2}
\newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
\newcommand{\tlin}[2]{#2 \subseteq \ms{lin}(#1)}
\newcommand{\ltlin}[3]{#3 \subseteq \ms{lin}(#1) \cap #2}
\newcommand{\thyp}[3]{#1: {#2}^{#3}}
\newcommand{\lhyp}[3]{#1[#2](#3)}
\newcommand{\llhyp}[3]{\lhyp{\lbl{#1}}{#2}{#3}}
\newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
\newcommand{\taff}{{\{\ms{a}\}}}
\newcommand{\trel}{{\{\ms{r}\}}}
\newcommand{\tint}{{\{\ms{a}, \ms{r}\}}}
\newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
\newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\issubst}[3]{#1: #2 \mapsto #3}
\newcommand{\lbsubst}[3]{#1: #2 \rightsquigarrow #3}
\newcommand{\exprletsubst}[2]{{#1};{#2}}
\newcommand{\stmtletsubst}[2]{{#1};{#2}}
\newcommand{\mhole}[1]{{#1}^?}
\newcommand{\lhole}[1]{?#1}
\newcommand{\mhasty}[6]{#1;#2 \vdash_{#3} #4: {#5}^{#6}}
\newcommand{\mhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\mlhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\tyhole}[5]{#1: #2 \mapsto_{#3} {#4}^{#5}}
\newcommand{\blkhole}[3]{#1: #2 \mapsto #3}
\newcommand{\cfghole}[3]{#1: #2 \mapsto #3}
\newcommand{\substctx}[2]{{#1}^{#2}}
\newcommand{\substlbs}[2]{{#1}^{#2}}
\newcommand{\restrictsubst}[2]{{#1}_{#2}}
\newcommand{\subsubst}[2]{#1 \subseteq #2}
\newcommand{\isrw}[3]{#1: #2 \mapsto #3}
\newcommand{\mbind}{\mathbin{{>}\hspace{-0.1em}{>}\hspace{-0.1em}{=}}}
% \newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}

@@

\TODO{cite \cite{rosen-gvn-1988} somewhere?}

@@

\TODO{clean up, and generalize to arbitrary premonoidal categories (w/ initial object? how to deal w/ coproducts?)}

@@

\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) is a pair \((\Pi, \cdot)\) where \(\Pi: M^\omega \to I\) is a map from streams of \(M\) to \(I\) and \(\cdot: M \to I \to I\) is a monoid action such that
  \(
    m \cdot \Pi \ell = \Pi (m: \ell) 
  \)
  where \(m:\ell\) denotes the stream formed by prefixing \(m\) to \(\ell\).
\end{definition}
Two important examples include:
\begin{itemize}
  \item For any monoid \(M\), the unique maps \(\Pi: (\nats \to M) \to \mb{1}\), \(\cdot: M \to \mb{1} \to \mb{1}\) form a stream action on \(\mb{1}\)
  \item The concatenation map \(\Pi: (A^*)^\omega \to A^\omega\) and the concatenation action \(\cdot: A^* \to A^{\leq \omega} \to A^{\leq \omega}\) form a stream action of the monoid of lists of \(A\), \(A^*\), on the set of potentially infinite sequences of \(A\), \(A^{\leq \omega}\).
\end{itemize}

\begin{definition}[Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Pi\;A = A \times M + I\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and \(\mu_A = [\lambda (a, m). m \cdot a, \ms{id}]\)
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with
  \[f^\dagger(a) = ...\]
\end{definition}

\begin{definition}[Nondeterministic Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\{\iota_0 (a, 1)\}\)
  and \(\mu_A = ...\)
  This is an Elgot monad, with
  \[f^\dagger(a) = ...\]
\end{definition}

\TODO{induces a substructural effectful category...}

\TODO{central = commutative, "commutes with infinity" (i.e. is absorbed by infinity)}

\TODO{relevant = deterministic}

\TODO{affine = no effect on trace}

@@

\TODO{text, \cite{sparky}}

\begin{definition}[Pomset] \
  \ \TODO{this}
\end{definition}

@@

We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.

\TODO{Name for ND trace monad}

@@

\TODO{\cite{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}

\TODO{\cite{schneider-phd}}

\TODO{\cite{chakravarty-functional-ssa-2003}}

\TODO{\cite{barthe-compcert-ssa-2014}}

\TODO{\cite{buchwald-verified-ssa-2016}}

\TODO{\cite{herklotz-gsa-2023}}

\TODO{\cite{pop-ssa-inout-2009}}

\TODO{\cite{garbuzov-structural-cfg-2018}}

\TODO{\cite{hua-explicit-ssa-2010}}

@@

\TODO{start appendix}
